Reasoning and Derivation of Monadic Programs

### 2019-04-16 paper/response.txt

We are very grateful for the helpful comments that all reviews make, and are
especially gladened by the expressions of appreciation. We will do our best to
address all comments in the final version, but focus here only on the main
issues and questions raised.

- Review B:

  About Dom not having a bind:

  There are two key ideas that make the emulation of local state by means
  of local state work:

    1. We impose a discipline that all uses of put are in a state-restoring
       fashion; no `wild' puts are allowed.

       => Without a bind, a carefully constructed disciplined Dom value cannot
          be extended in a sequential (i.e., bind-) fashion with a wild put.

    2. The state at the end of a branch is transferred to (the beginning of)
       the next branch.

       => Without a bind, we know where a branch is at its end and can build in
          the transfer of the state to the next branch. With a bind, we would
	  not know when we are at the end of the branch and would have to defer
          the transfer, leaving room for the possibility that an entirely different
          state value is passed to the next branch.:w

  We will articulate this better. You make a good point here about moving the concrete Dom 
  from the appendix to the main text. Then we will be able to constrast with ListT-done
  right applied to the state monad, which suffers from the above two problems.

- Review C:

  Novelty wrt "Just do It":
  TODO

  The use of Haskell:
  We feel that the use of Haskell for programming and reasoning with monads is justified given
  the extensive existing body of literature that does just that. Nevertheless,
  we appreciate your pointing out that the accessibility of the paper will be improved when we 
  add more explanation of concepts like hylomorphisms and scanl.

  Benefits of Monads:
  A big benefit of monads is that it provides a pureful functional model of
  effects. This has become an actual practice in Haskell and is a good model for
  what happens in languages with algebraic effects & handlers. Yet, we should
  emphasize that the focus of our paper is reasoning about monadic programs, not
  to justify the use of monads. We are not aware of any other formalism used
  to tackle the same problem for other effect models. Even if it would exist,
  we believe that our work is still relevant because it directly applies to 
  the above settings where monads are used.

  (Non-)Modularity:
  There is modularity and non-modularity at different levels, which should not
  be mixed up.

    As we require a particular interaction between state and non-determinism,
    namely the local state semantics, by definition we do not want to reason about
    the two effects in isolation.

    Reasoning about local state happens modularly, and not for a whole program 
    at once. We use the local state laws for that.

    Enabling this local reasoning about local state, when implemented by means of
    global state, requires a global discipline of modifying state only in a 
    restoring-upon-backtracking fashion. We carry out this proof once, which involves
    reasoning about all disciplined whole programs, so that we can then reason
    locally about all specific programs.

    Syntax and (Algebraic) Effect Handlers Approach:
    (Not "event handlers")

    TODO

- Review D:

   TODO

- A1:

  We are grateful for a better reference of Prolog's 4-port model.

### 2019-04-24 mpc2019/paper/der-monad.tex
bibliography works now, plus various cosmetic fixes

% %% Some recommended packages.
% \usepackage{booktabs}   %% For formal tables:
%                         %% http://ctan.org/pkg/booktabs
% \usepackage{subcaption} %% For complex figures with subfigures/subcaptions
%                         %% http://ctan.org/pkg/subcaption

@@

% % \makeatletter\if@ACM@journal\makeatother
% % %% Journal information (used by PACMPL format)
% % %% Supplied to authors by publisher for camera-ready submission
% % \acmJournal{PACMPL}
% % \acmVolume{1}
% % \acmNumber{1}
% % \acmArticle{1}
% % \acmYear{2017}
% % \acmMonth{1}
% % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
% % \startPage{1}
% % \else
% \makeatother
% 
% %% Journal information
% %% Supplied to authors by publisher for camera-ready submission;
% %% use defaults for review submission.
% \acmJournal{PACMPL}
% \acmVolume{1}
% \acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
% \acmArticle{1}
% \acmYear{2019}
% \acmMonth{1}
% \acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
% \startPage{1}
% 
% %% Copyright information
% %% Supplied to authors (based on authors' rights management selection;
% %% see authors.acm.org) by publisher for camera-ready submission
% \setcopyright{none}             %% For review submission
% %\setcopyright{acmcopyright}
% %\setcopyright{acmlicensed}
% %\setcopyright{rightsretained}
% %\copyrightyear{2017}           %% If different from \acmYear

%% Bibliography style
% \bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
%\citestyle{acmauthoryear}   %% For author/year citations

@@

%% Title information
\title{Reasoning and Derivation of Monadic Programs}
\titlerunning{Deriving Monadic Programs}
%% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
\subtitle{A Case Study of Non-determinism and State}
%% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\institute{Institute of Information Science, Academia Sinica
    \and Department of Computer Science, KU Leuven
    }
\author{Shin-Cheng Mu \inst{1} \and Tom Schrijvers \inst{2} \and Koen Pauwels \inst{2}}
%\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
% TODO
% \affiliation{
% %  \position{Position1}
%   \department{Institute of Information Science}              %% \department is recommended
%   \institution{Academia Sinica}            %% \institution is required
% %  \streetaddress{Street1 Address1}
% %  \city{City1}
% %  \state{State1}
% %  \postcode{Post-Code1}
%   \country{Taiwan}
% }
%\email{scm@iis.sinica.edu.tw}          %% \email is recommended

% \author{Tom Schrijvers}
%\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
% \affiliation{
% %  \position{Position1}
%   \department{Department of Computer Science}              %% \department is recommended
%   \institution{KU Leuven}            %% \institution is required
% %  \streetaddress{Street1 Address1}
% %  \city{City1}
% %  \state{State1}
% %  \postcode{Post-Code1}
%   \country{Belgium}
% }
%\email{tom.schrijvers@cs.kuleuven.be} %% \email is recommended

% \author{Koen Pauwels}
%\authornote{with author1 note}          %% \authornote is optional;
                                         %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
% \affiliation{
% %  \position{Position1}
%   \department{Department of Computer Science}              %% \department is recommended
%   \institution{KU Leuven}            %% \institution is required
% %  \streetaddress{Street1 Address1}
% %  \city{City1}
% %  \state{State1}
% %  \postcode{Post-Code1}
%   \country{Belgium}
% }
%\email{koen.pauwels@student.kuleuven.be} %% \email is recommended

@@

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
% \begin{CCSXML}
% <ccs2012>
% <concept>
% <concept_id>10003752.10010124.10010138</concept_id>
% <concept_desc>Theory of computation~Program reasoning</concept_desc>
% <concept_significance>500</concept_significance>
% </concept>
% </ccs2012>
% \end{CCSXML}

% \ccsdesc[500]{Theory of computation~Program reasoning}
%% End of generated code

%% Keywords
%% comma separated list
% \keywords{monad, effects, program derivation, equational reasoning}  %% \keywords is optional

@@

It is misleading if one says that functional programming does not allow side effects. In fact, even a purely functional language may allow a variety of side effects --- in a rigorous, mathematically manageable manner. Since the introduction of {\em monads} into the functional programming community~\cite{Moggi:89:Computational, Wadler:92:Monads}, it has become the main framework in which effects are modelled. Various monads were developed for different effects, from general ones such as IO, state, non-determinism, exception, continuation, environment passing, to specific purposes such as parsing. Numerous research were also devoted to producing practical monadic programs.

@@

When mixed with state, one consequence of \eqref{eq:mplus-bind-dist} is that \ensuremath{\Varid{get}\mathrel{\hstretch{0.7}{>\!\!>\!\!=}}(\lambda \Varid{s}\to \Varid{f}_{1}\;\Varid{s}\mathbin{\talloblong}\Varid{f}_{2}\;\Varid{s})\mathrel{=}(\Varid{get}\mathrel{\hstretch{0.7}{>\!\!>\!\!=}}\Varid{f}_{1}\mathbin{\talloblong}\Varid{get}\mathrel{\hstretch{0.7}{>\!\!>\!\!=}}\Varid{f}_{2})}. That is, \ensuremath{\Varid{f}_{1}} and \ensuremath{\Varid{f}_{2}} get the same state regardless of whether \ensuremath{\Varid{get}} is performed outside or inside the non-deterministic branch.
Similarly, \eqref{eq:mzero-bind-zero} implies \ensuremath{\Varid{put}\;\Varid{s}\mathbin{\hstretch{0.7}{>\!\!>}}\emptyset\mathrel{=}\emptyset} --- when a program fails, the changes it performed on the state can be discarded.
These requirements imply that each non-deterministic branch has its own copy of the state.
Therefore, we will refer to \eqref{eq:mplus-bind-dist} and \eqref{eq:mzero-bind-zero} as \emph{local state laws} in this paper --- even though they do not explicitly mention state operators at all!
One monad satisfying the laws is \ensuremath{\Conid{M}_{\{\mskip1.5mu \Conid{N},\Conid{S}\;\Varid{s}\mskip1.5mu\}}\;\Varid{a}\mathrel{=}\Varid{s}\to [\mskip1.5mu (\Varid{a},\Varid{s})\mskip1.5mu]}, which is the same monad one gets by \ensuremath{\Conid{StateT}\;\Varid{s}\;(\Conid{ListT}\;\Conid{Identity})} in the Monad Transformer Library~\cite{MTL:14}.
With effect handling~\cite{Wu:14:Effect, KiselyovIshii:15:Freer}, the monad meets the requirements if we run the handler for state before that for list.

@@

\paragraph{Note} We briefly justify proofs by induction on the syntax tree.
Finite monadic programs can be represented by the free monad constructed out of \ensuremath{\Varid{return}} and the effect operators, which can be represented by an inductively defined data structure, and interpreted by effect handlers~\cite{Kiselyov:13:Extensible, KiselyovIshii:15:Freer}.
When we say two programs \ensuremath{\Varid{m}_{1}} and \ensuremath{\Varid{m}_{2}} are equal, we mean that they have the same denotation when interpreted by the effect handlers of the corresponding effects, for example, \ensuremath{\Varid{hdNondet}\;(\Varid{hdState}\;\Varid{s}\;\Varid{m}_{1})\mathrel{=}\Varid{hdNondet}\;(\Varid{hdState}\;\Varid{s}\;\Varid{m}_{2})}, where \ensuremath{\Varid{hdNondet}} and \ensuremath{\Varid{hdState}} are respectively handlers for nondeterminism and state.
Such equality can be proved by induction on some sub-expression in \ensuremath{\Varid{m}_{1}} or \ensuremath{\Varid{m}_{2}}, which are treated like any inductively defined data structure.
A more complete treatment is a work in progress, which cannot be fully covered in this paper.
({\em End of Note})

@@

\koen{TODO: LLNCS docs recommend not using vspace command}
\vspace{1em}
Theorem \ref{thm:hylo-fusion} does not rely on the \emph{local state laws} \eqref{eq:mplus-bind-dist} and \eqref{eq:mzero-bind-zero}, and does not put restriction on \ensuremath{\epsilon}.
To apply the theorem to our particular case, we have to show that its preconditions hold for our particular \ensuremath{(\odot)} ---
for that we will need \eqref{eq:mplus-bind-dist} and \eqref{eq:mzero-bind-zero}. In the lemma below we slightly generalise \ensuremath{(\odot)} in Theorem \ref{lma:foldr-guard-fusion}:
\begin{lemma} Assuming that state and non-determinism commute, and \ensuremath{\Varid{m}\mathrel{\hstretch{0.7}{>\!\!>\!\!=}}\emptyset\mathrel{=}\emptyset}. Given \ensuremath{\Varid{p}\mathbin{::}\Varid{a}\to \Varid{s}\to \Conid{Bool}}, \ensuremath{\Varid{next}\mathbin{::}\Varid{a}\to \Varid{s}\to \Varid{s}}, \ensuremath{\Varid{res}\mathbin{::}\Varid{a}\to \Varid{b}\to \Varid{b}}, define \ensuremath{(\odot)\mathbin{::}\Varid{a}\to \Conid{M}_{\epsilon}\;\Varid{b}\to \Conid{M}_{\epsilon}\;\Varid{b}} with \ensuremath{\{\mskip1.5mu \Conid{N},\Conid{S}\;\Varid{s}\mskip1.5mu\}\mathbin{\subseteq}\epsilon}:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\Varid{x}\mathbin{\odot}\Varid{m}\mathrel{=}{}\<[17]%
\>[17]{}\Varid{get}\mathrel{\hstretch{0.7}{>\!\!>\!\!=}}\lambda \Varid{st}\to \Varid{guard}\;(\Varid{p}\;\Varid{x}\;\Varid{st})\mathbin{\hstretch{0.7}{>\!\!>}}{}\<[E]%
\\
\>[17]{}\Varid{put}\;(\Varid{next}\;\Varid{x}\;\Varid{st})\mathbin{\hstretch{0.7}{>\!\!>}}(\Varid{res}\;\Varid{x}\mathrel{\raisebox{0.5\depth}{\scaleobj{0.5}{\langle}} \scaleobj{0.8}{\$} \raisebox{0.5\depth}{\scaleobj{0.5}{\rangle}}}\Varid{m})~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
We have \ensuremath{\Varid{x}\mathbin{\odot}\Varid{m}\mathrel{=}\Varid{m}\mathrel{\hstretch{0.7}{>\!\!>\!\!=}}((\Varid{x}\mathbin{\odot})\mathbin{\cdot}\Varid{return})}.
\end{lemma}
\begin{proof}
Routine, using commutativity of state and non-determinism.
\end{proof}

@@

@@

\>[11]{}\mathrel{=}\Varid{unfoldM}\;\Varid{null}\;\Varid{select}\;[\mskip1.5mu \mathrm{0}\mathinner{\ldotp\ldotp}\Varid{n}\mathbin{-}\mathrm{1}\mskip1.5mu]\mathrel{\hstretch{0.7}{>\!\!>\!\!=}}\Varid{filt}\;(\Varid{all}\;\Varid{ok}\mathbin{\cdot}\Varid{scanl}_{+}\;(\oplus)\;(\mathrm{0},[\mskip1.5mu \mskip1.5mu],[\mskip1.5mu \mskip1.5mu]))~~,{}\<[E]%

@@

Non-deterministic monads with a global state, however, is rather tricky.
One might believe that \ensuremath{\Conid{M}\;\Varid{a}\mathrel{=}\Varid{s}\to ([\mskip1.5mu \Varid{a}\mskip1.5mu],\Varid{s})} is a natural implementation of such a monad.
The usual, naive implementation of \ensuremath{(\mathrel{\hstretch{0.7}{>\!\!>\!\!=}})} using this representation, however, does not satisfy left-distributivity \eqref{eq:bind-mplus-dist}, violates monad laws, and is therefore not even a monad.
%See Section \ref{sec:conclusion} for previous work on construction of a correct monad.
\footnote{
The type \ensuremath{\Conid{ListT}\;(\Conid{State}\;\Varid{s})} generated using the now standard Monad Transformer Library~\cite{MTL:14} expands to essentially the same implementation, and is flawed in the same way.
More careful implementations of \ensuremath{\Conid{ListT}}, which does satisfy \eqref{eq:bind-mplus-dist} and the monad laws, have been proposed~\cite{Gale:07:ListT, Volkov:14:list-t}.
Effect handlers (e.g. Wu~\cite{Wu:14:Effect} and Kiselyov and Ishii~\cite{KiselyovIshii:15:Freer}) do produce correct implementations if we run the handler for non-determinism before that of state.
}

@@

\small

@@

@@

\>[7]{}\scaleobj{0.8}{\langle}\Varid{put}\scaleobj{0.8}{\rangle}\;\Varid{x}\;(\scaleobj{0.8}{\langle}\Varid{put}\scaleobj{0.8}{\rangle}\;\Varid{y}\;(\scaleobj{0.8}{\langle}\Varid{ret}\scaleobj{0.8}{\rangle}\;\Varid{w}))\mathbin{\scaleobj{0.8}{\langle[\!]\rangle}}\scaleobj{0.8}{\langle}\Varid{put}\scaleobj{0.8}{\rangle}\;\Varid{x}\;(\scaleobj{0.8}{\langle}\Varid{get}\scaleobj{0.8}{\rangle}\;(\lambda \Varid{s}\to \scaleobj{0.8}{\langle}\Varid{put}\scaleobj{0.8}{\rangle}\;\Varid{y}\;(\scaleobj{0.8}{\langle}\Varid{ret}\scaleobj{0.8}{\rangle}\;\Varid{s}))){}\<[E]%

@@

The proof for both the \ensuremath{\Varid{m}\mathrel{=}\Conid{Get}\;\Varid{k}} and \ensuremath{\Varid{m}\mathrel{=}\Conid{Put}\;\Varid{s}\;\Varid{m'}} cases requires that \ensuremath{\Conid{Get}}
distributes
over \ensuremath{(\talloblong)} at the top-level of a global-state program if the left branch is
state restoring.
\begin{align}
  % get_or_trans
\ensuremath{\Varid{run}\;(\Conid{Get}\;(\lambda \Varid{s}\to \Varid{trans}\;(\Varid{m}_{1}\;\Varid{s})\mathbin{\talloblong}(\Varid{m}_{2}\;\Varid{s})))} = \ensuremath{\Varid{run}\;(\Conid{Get}\;(\lambda \Varid{s}\to \Varid{trans}\;(\Varid{m}_{1}\;\Varid{s}))\mathbin{\talloblong}\Conid{Get}\;\Varid{m}_{2})} \label{eq:get-ret-mplus-g}\mbox{~~.} \checkmark
\end{align}

@@

We then define two translation functions from \ensuremath{\Conid{Prog\char95 m}\;\Varid{a}} to \ensuremath{\Conid{Prog}\;\Varid{a}},
which both replace \ensuremath{\Conid{Put}}s with \ensuremath{\Varid{put}_{\scaleobj{0.7}{\sf R}}}s along the way, like the regular
\ensuremath{\Varid{trans}} function.
The first replaces each \ensuremath{\Conid{Modify\char95 R}} in the program by a direct analogue of the
definition given above, while the second replaces it by
\ensuremath{\Conid{Get}\;(\lambda \Varid{s}\to \Conid{Put}\;(\Varid{next}\;\Varid{s})\;(\Varid{trans\char95 2}\;\Varid{p}))}:

>>>>

@@

@@

%% Title information
\title{Reasoning and Derivation of Monadic Programs}
\titlerunning{Deriving Monadic Programs}
\subtitle{A Case Study of Non-determinism and State}

\institute{Institute of Information Science, Academia Sinica, Taiwan, \email{scm@iis.sinica.edu.tw}
    \and Department of Computer Science, KU Leuven, Belgium, \email{first.last@cs.kuleuven.be}
    }
\author{Shin-Cheng Mu \inst{1} \and Tom Schrijvers \inst{2} \and Koen Pauwels \inst{2}}

@@

\keywords{monads \and effects \and program derivation \and equational reasoning \and nondeterminism \and state}

@@

It is misleading if one says that functional programming does not allow side effects. In fact, even a purely functional language may allow a variety of side effects --- in a rigorous, mathematically manageable manner. Since the introduction of {\em monads} into the functional programming community~\cite{Moggi:89:Computational,Wadler:92:Monads}, it has become the main framework in which effects are modelled. Various monads were developed for different effects, from general ones such as IO, state, non-determinism, exception, continuation, environment passing, to specific purposes such as parsing. Numerous research were also devoted to producing practical monadic programs.

@@

When mixed with state, one consequence of \eqref{eq:mplus-bind-dist} is that \ensuremath{\Varid{get}\mathrel{\hstretch{0.7}{>\!\!>\!\!=}}(\lambda \Varid{s}\to \Varid{f}_{1}\;\Varid{s}\mathbin{\talloblong}\Varid{f}_{2}\;\Varid{s})\mathrel{=}(\Varid{get}\mathrel{\hstretch{0.7}{>\!\!>\!\!=}}\Varid{f}_{1}\mathbin{\talloblong}\Varid{get}\mathrel{\hstretch{0.7}{>\!\!>\!\!=}}\Varid{f}_{2})}. That is, \ensuremath{\Varid{f}_{1}} and \ensuremath{\Varid{f}_{2}} get the same state regardless of whether \ensuremath{\Varid{get}} is performed outside or inside the non-deterministic branch.
Similarly, \eqref{eq:mzero-bind-zero} implies \ensuremath{\Varid{put}\;\Varid{s}\mathbin{\hstretch{0.7}{>\!\!>}}\emptyset\mathrel{=}\emptyset} --- when a program fails, the changes it performed on the state can be discarded.
These requirements imply that each non-deterministic branch has its own copy of the state.
Therefore, we will refer to \eqref{eq:mplus-bind-dist} and \eqref{eq:mzero-bind-zero} as \emph{local state laws} in this paper --- even though they do not explicitly mention state operators at all!
One monad satisfying the laws is \ensuremath{\Conid{M}_{\{\mskip1.5mu \Conid{N},\Conid{S}\;\Varid{s}\mskip1.5mu\}}\;\Varid{a}\mathrel{=}\Varid{s}\to [\mskip1.5mu (\Varid{a},\Varid{s})\mskip1.5mu]}, which is the same monad one gets by \ensuremath{\Conid{StateT}\;\Varid{s}\;(\Conid{ListT}\;\Conid{Identity})} in the Monad Transformer Library~\cite{MTL:14}.
With effect handling~\cite{Wu:14:Effect,KiselyovIshii:15:Freer}, the monad meets the requirements if we run the handler for state before that for list.

@@

\paragraph{Note} We briefly justify proofs by induction on the syntax tree.
Finite monadic programs can be represented by the free monad constructed out of \ensuremath{\Varid{return}} and the effect operators, which can be represented by an inductively defined data structure, and interpreted by effect handlers~\cite{Kiselyov:13:Extensible,KiselyovIshii:15:Freer}.
When we say two programs \ensuremath{\Varid{m}_{1}} and \ensuremath{\Varid{m}_{2}} are equal, we mean that they have the same denotation when interpreted by the effect handlers of the corresponding effects, for example, \ensuremath{\Varid{hdNondet}\;(\Varid{hdState}\;\Varid{s}\;\Varid{m}_{1})\mathrel{=}\Varid{hdNondet}\;(\Varid{hdState}\;\Varid{s}\;\Varid{m}_{2})}, where \ensuremath{\Varid{hdNondet}} and \ensuremath{\Varid{hdState}} are respectively handlers for nondeterminism and state.
Such equality can be proved by induction on some sub-expression in \ensuremath{\Varid{m}_{1}} or \ensuremath{\Varid{m}_{2}}, which are treated like any inductively defined data structure.
A more complete treatment is a work in progress, which cannot be fully covered in this paper.
({\em End of Note})

@@

% \vspace{1em} % Koen: LLNCS docs ask to avoid vspace, can be allowed in "exceptional cases"
Theorem \ref{thm:hylo-fusion} does not rely on the \emph{local state laws} \eqref{eq:mplus-bind-dist} and \eqref{eq:mzero-bind-zero}, and does not put restriction on \ensuremath{\epsilon}.
To apply the theorem to our particular case, we have to show that its preconditions hold for our particular \ensuremath{(\odot)} ---
for that we will need \eqref{eq:mplus-bind-dist} and \eqref{eq:mzero-bind-zero}. In the lemma below we slightly generalise \ensuremath{(\odot)} in Theorem \ref{lma:foldr-guard-fusion}:
\begin{lemma} Assuming that state and non-determinism commute, and \ensuremath{\Varid{m}\mathrel{\hstretch{0.7}{>\!\!>\!\!=}}\emptyset\mathrel{=}\emptyset}. Given \ensuremath{\Varid{p}\mathbin{::}\Varid{a}\to \Varid{s}\to \Conid{Bool}}, \ensuremath{\Varid{next}\mathbin{::}\Varid{a}\to \Varid{s}\to \Varid{s}}, \ensuremath{\Varid{res}\mathbin{::}\Varid{a}\to \Varid{b}\to \Varid{b}}, define \ensuremath{(\odot)\mathbin{::}\Varid{a}\to \Conid{M}_{\epsilon}\;\Varid{b}\to \Conid{M}_{\epsilon}\;\Varid{b}} with \ensuremath{\{\mskip1.5mu \Conid{N},\Conid{S}\;\Varid{s}\mskip1.5mu\}\mathbin{\subseteq}\epsilon}:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\Varid{x}\mathbin{\odot}\Varid{m}\mathrel{=}{}\<[17]%
\>[17]{}\Varid{get}\mathrel{\hstretch{0.7}{>\!\!>\!\!=}}\lambda \Varid{st}\to \Varid{guard}\;(\Varid{p}\;\Varid{x}\;\Varid{st})\mathbin{\hstretch{0.7}{>\!\!>}}{}\<[E]%
\\
\>[17]{}\Varid{put}\;(\Varid{next}\;\Varid{x}\;\Varid{st})\mathbin{\hstretch{0.7}{>\!\!>}}(\Varid{res}\;\Varid{x}\mathrel{\raisebox{0.5\depth}{\scaleobj{0.5}{\langle}} \scaleobj{0.8}{\$} \raisebox{0.5\depth}{\scaleobj{0.5}{\rangle}}}\Varid{m})~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
We have \ensuremath{\Varid{x}\mathbin{\odot}\Varid{m}\mathrel{=}\Varid{m}\mathrel{\hstretch{0.7}{>\!\!>\!\!=}}((\Varid{x}\mathbin{\odot})\mathbin{\cdot}\Varid{return})}.
\end{lemma}
\begin{proof}
Routine, using commutativity of state and non-determinism.
\end{proof}

@@

\column{13}{@{}>{\hspre}l<{\hspost}@{}}%

@@

\>[11]{}\mathrel{=}\Varid{unfoldM}\;\Varid{null}\;\Varid{select}\;[\mskip1.5mu \mathrm{0}\mathinner{\ldotp\ldotp}\Varid{n}\mathbin{-}\mathrm{1}\mskip1.5mu]\mathrel{\hstretch{0.7}{>\!\!>\!\!=}}{}\<[E]%
\>[11]{}\hsindent{2}{}\<[13]%
\>[13]{}\Varid{filt}\;(\Varid{all}\;\Varid{ok}\mathbin{\cdot}\Varid{scanl}_{+}\;(\oplus)\;(\mathrm{0},[\mskip1.5mu \mskip1.5mu],[\mskip1.5mu \mskip1.5mu]))~~,{}\<[E]%

@@

Non-deterministic monads with a global state, however, is rather tricky.
One might believe that \ensuremath{\Conid{M}\;\Varid{a}\mathrel{=}\Varid{s}\to ([\mskip1.5mu \Varid{a}\mskip1.5mu],\Varid{s})} is a natural implementation of such a monad.
The usual, naive implementation of \ensuremath{(\mathrel{\hstretch{0.7}{>\!\!>\!\!=}})} using this representation, however, does not satisfy left-distributivity \eqref{eq:bind-mplus-dist}, violates monad laws, and is therefore not even a monad.
%See Section \ref{sec:conclusion} for previous work on construction of a correct monad.
\footnote{
The type \ensuremath{\Conid{ListT}\;(\Conid{State}\;\Varid{s})} generated using the now standard Monad Transformer Library~\cite{MTL:14} expands to essentially the same implementation, and is flawed in the same way.
More careful implementations of \ensuremath{\Conid{ListT}}, which does satisfy \eqref{eq:bind-mplus-dist} and the monad laws, have been proposed~\cite{Gale:07:ListT,Volkov:14:list-t}.
Effect handlers (e.g. Wu~\cite{Wu:14:Effect} and Kiselyov and Ishii~\cite{KiselyovIshii:15:Freer}) do produce correct implementations if we run the handler for non-determinism before that of state.
}

@@

@@

  \tiny

@@

\>[7]{}\scaleobj{0.8}{\langle}\Varid{put}\scaleobj{0.8}{\rangle}\;\Varid{x}\;(\scaleobj{0.8}{\langle}\Varid{put}\scaleobj{0.8}{\rangle}\;\Varid{y}\;(\scaleobj{0.8}{\langle}\Varid{ret}\scaleobj{0.8}{\rangle}\;\Varid{w})){}\<[E]%
\>[9]{}\mathbin{\scaleobj{0.8}{\langle[\!]\rangle}}\scaleobj{0.8}{\langle}\Varid{put}\scaleobj{0.8}{\rangle}\;\Varid{x}\;(\scaleobj{0.8}{\langle}\Varid{get}\scaleobj{0.8}{\rangle}\;(\lambda \Varid{s}\to \scaleobj{0.8}{\langle}\Varid{put}\scaleobj{0.8}{\rangle}\;\Varid{y}\;(\scaleobj{0.8}{\langle}\Varid{ret}\scaleobj{0.8}{\rangle}\;\Varid{s}))){}\<[E]%

@@

The proof for both the \ensuremath{\Varid{m}\mathrel{=}\Conid{Get}\;\Varid{k}} and \ensuremath{\Varid{m}\mathrel{=}\Conid{Put}\;\Varid{s}\;\Varid{m'}} cases requires that \ensuremath{\Conid{Get}}
distributes
over \ensuremath{(\talloblong)} at the top-level of a global-state program if the left branch is
state restoring.
\begin{align}
  % get_or_trans
\ensuremath{\Varid{run}\;(\Conid{Get}\;(\lambda \Varid{s}\to \Varid{trans}\;(\Varid{m}_{1}\;\Varid{s})\mathbin{\talloblong}(\Varid{m}_{2}\;\Varid{s})))} = \\
\ensuremath{\Varid{run}\;(\Conid{Get}\;(\lambda \Varid{s}\to \Varid{trans}\;(\Varid{m}_{1}\;\Varid{s}))\mathbin{\talloblong}\Conid{Get}\;\Varid{m}_{2})} \label{eq:get-ret-mplus-g}\mbox{~~.} \checkmark
\end{align}

@@

We then define two translation functions from \ensuremath{\Conid{Prog\char95 m}\;\Varid{a}} to \ensuremath{\Conid{Prog}\;\Varid{a}},
which both replace \ensuremath{\Conid{Put}}s with \ensuremath{\Varid{put}_{\scaleobj{0.7}{\sf R}}}s along the way, like the regular
\ensuremath{\Varid{trans}} function.
The first replaces each \ensuremath{\Conid{Modify\char95 R}} in the program by a direct analogue of the
definition given above, while the second replaces it by
\ensuremath{\Conid{Get}\;(\lambda \Varid{s}\to \Conid{Put}\;(\Varid{next}\;\Varid{s})\;(\Varid{trans}_{\scaleobj{0.7}{2}}\;\Varid{p}))}:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
\column{36}{@{}>{\hspre}l<{\hspost}@{}}%
\column{45}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[4]{}\Varid{trans}_{\scaleobj{0.7}{1}}\mathbin{::}\Conid{Prog\char95 m}\;\Varid{a}\to \Conid{Prog}\;\Varid{a}{}\<[E]%
\\
\>[4]{}\mathbin{...}{}\<[E]%
\\
\>[4]{}\Varid{trans}_{\scaleobj{0.7}{1}}\;(\Conid{Modify\char95 R}\;\Varid{next}\;\Varid{prev}\;\Varid{p}){}\<[36]%
\>[36]{}\mathrel{=}{}\<[45]%
\>[45]{}\Conid{Get}\;(\lambda \Varid{s}\to \Conid{Put}\;(\Varid{next}\;\Varid{s})\;(\Varid{trans}_{\scaleobj{0.7}{1}}\;\Varid{p}){}\<[E]%
\\
\>[36]{}\mathbin{\talloblong}{}\<[45]%
\>[45]{}\Conid{Get}\;(\lambda \Varid{t}\to \Conid{Put}\;(\Varid{prev}\;\Varid{t})\;\emptyset)){}\<[E]%
\\[\blanklineskip]%
\>[4]{}\Varid{trans}_{\scaleobj{0.7}{2}}\mathbin{::}\Conid{Prog\char95 m}\;\Varid{a}\to \Conid{Prog}\;\Varid{a}{}\<[E]%
\\
\>[4]{}\mathbin{...}{}\<[E]%
\\
\>[4]{}\Varid{trans}_{\scaleobj{0.7}{2}}\;(\Conid{Modify\char95 R}\;\Varid{next}\;\Varid{prev}\;\Varid{p}){}\<[36]%
\>[36]{}\mathrel{=}\Conid{Get}\;(\lambda \Varid{s}\to \Varid{put}_{\scaleobj{0.7}{\sf R}}\;(\Varid{next}\;\Varid{s})\;(\Varid{trans}_{\scaleobj{0.7}{2}}\;\Varid{p})){}\<[E]%
\\
\>[4]{}\hsindent{2}{}\<[6]%
\>[6]{}\mathbf{where}\;\Varid{put}_{\scaleobj{0.7}{\sf R}}\;\Varid{s}\;\Varid{p}\mathrel{=}\Conid{Get}\;(\lambda \Varid{t}\to \Conid{Put}\;\Varid{s}\;\Varid{p}\mathbin{\talloblong}\Conid{Put}\;\Varid{t}\;\emptyset){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

It is clear that \ensuremath{\Varid{trans}_{\scaleobj{0.7}{2}}\;\Varid{p}} is the exact same program as \ensuremath{\Varid{trans}\;\Varid{p'}}, where \ensuremath{\Varid{p'}}
is \ensuremath{\Varid{p}} but with each \ensuremath{\Conid{ModifyR}\;\Varid{next}\;\Varid{prev}\;\Varid{p}} replaced by \ensuremath{\Conid{Get}\;(\lambda \Varid{s}\to \Conid{Put}\;(\Varid{next}\;\Varid{s})\;\Varid{p})}.
